---
title: "ggproto API"
author: "Nicholas Spyrison"
date: '`r format(Sys.Date())`'
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggproto API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---
<!-- #Example vignette:
https://github.com/njtierney/naniar/blob/master/vignettes/getting-started-w-naniar.Rmd -->
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,   # code
  include    = TRUE,   # plots
  results    = "show", # text: "hide", "show"
  eval       = TRUE,   # chunk
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  collapse   = TRUE,
  comment    = "#>",
  fig.height = 4,
  fig.width  = 6,
  fig.align  = "center",
  cache      = FALSE
)
```


We are moving to a new '`ggproto`' API for constructing animated tours from `{ggplot2}` objects. This interface should feel more comfortable to people already familiar with `{ggplot2}`. Ggproto is the parent class of geoms, layers, and some other plotting elements.

Proto objects (or lists of them) can be assigned to a variable without `ggplot()`, they store _unevaluated_ code that will be used to plot. Let's see what this looks like in ggplot2 before we delve into tour animations.

```{r}
library("ggplot2")
library("magrittr")

## A ggproto:
gp <- geom_point()
class(gp)

## A list of ggplot elements, a 'head-less' ggplot call
gg_ls <- list(
  gp,
  geom_smooth(method = "loess", formula = y ~ x),
  ggtitle("proto_* functions return lists of geoms_* functions.", "These lists can be stored, extended, and added to ggplot(). \n We use this to include the animation of ggplots."),
  facet_grid(cols = vars(Tree))
)
lapply(gg_ls, class)

## ggplot call, without geoms, a 'body-less' ggplot call
gghead <- ggplot(Orange, aes(age, circumference, color = Tree))

## Evaluate together
gghead +
  gg_ls
```

In the `ggproto` API we use this feature to create `proto_*()` functions, the counter-parts of `geom_*()` functions. These proto functions are used in composition with `ggtour()`, which replaces `ggplot()`, to create animated tours. In adopting this additive, composition approach we aim to maximize the flexibility for the end-users to customize tours while keeping the door open for extension of the development of further `proto` functions.

```{r, echo = FALSE}
data.frame(
  object = c("head", "body", "render"),
  ggplot2 = c("ggplot()", "geom_*()", "NA"),
  `ggproto api` = c("ggtour()", "proto_*()", "animate_*()"),
  `previous api` = c("play_manual_tour()", "play_manual_tour()", "play_manual_tour(render_*())")
) %>%
  knitr::kable()
```

`play_manual_tour()` abstracted away complexity, but it was becoming too big, and consuming workflow which a function should venue too far into. This made it too bloated to allow for some flexibility, but hard to understand all of the arguments. By paring back to `proto_*` can keep details where they are more relevant without becoming overbearing in one monster function.

## Why 'proto_\*' instead of new 'geom_\*'

`{spinifex} proto_*` functions return a list of several `geom_*` functions, that were designed to facilitate animation across many projection bases. We abstract away of work and complexity that comes with creating and animating tours, but this comes at the price of flexibility. For instance, protos do not know how to deal with facets and hard-code the number and type of geoms which would otherwise become very burdensome to specify all the specifics of.

## Application

### Manual tour
```{r}
library("tourr")
library("spinifex")

## Scale our numeric data
dat <- scale_sd(tourr::flea[, 1:6])
## Use species as a class to set color and shape with
clas <- tourr::flea$species

## Manual tour, manipulating the contribution of a selected variable 
bas <- basis_pca(dat) ## Start basis
mv <- manip_var_of(bas) ## Number of the variable to manipulate
mt_path <- manual_tour(bas, manip_var = mv) ## Tour path

## Create a static ggplot2 plot with all frames of the tour
ggt <- ggtour(mt_path, dat, angle = .1) +
  proto_basis() +
  proto_point(aes_args = list(color = clas, shape = clas),
              identity_args = list(size = 1.5))
```
```{r, eval=FALSE}
## Animate the tour with an animate_* function.
animate_plotly(ggt, fps = 5) ## As a plotly .html widget.
#animate_gganimate(ggt, fps = 8, rewind = TRUE) ## As a gganimate .gif
```

### Tours from `{tourr}`
```{r}
## (Quietly create) a grand tour, projecting through randomly selected bases
.mute <- utils::capture.output(
  gt_path <- tourr::save_history(dat, grand_tour(), max_bases = 3)
)

## Static ggplot of all frames in the tour
ggt <- ggtour(gt_path, dat, angle = .1) + ## Include geodesic interpolation angle between the selected bases.
  proto_basis(position = "topright") +
  proto_point(list(color = clas, shape = clas))
```
```{r, eval=FALSE}
## Animate
animate_plotly(ggt)
animate_gganimate(ggt)
```

### 1D tours
```{r}
## (Quietly create) a 1d guided tour, optimizing the projection space for the holes() function
guided_path <- save_history(dat, guided_tour(holes(), d = 1))

## Static ggplot of all frames in the tour
ggt <- ggtour(guided_path, dat, angle = .1) + ## Include geodesic interpolation angle between the selected bases.
  proto_basis1d() +
  proto_density(list(fill = clas), density_position = "stack")
## Note that proto_density(density_position = "stack") does not work with animate_plotly().
```
```{r, eval=FALSE}
## Animate
animate_gganimate(ggt)
```

### interation with ggplot2 functions

because the output of `ggtour() + proto_*` is a ggplot, users are encouraged to theme and setting functions as well.

```{r}
ggt <- ggt +
  theme_bw() +
  ggtitle("My Tour animation") +
  labs(x = "Projection Y1", y = "density")
```
```{r, eval=FALSE}
animate_gganimate(ggt)
```


## Proto functions

```{r, echo = FALSE}
data.frame(
  `proto functions` =
    c("ggtour", "proto_point", "proto_text", "proto_hex", "proto_origin/1d", "proto_density", "proto_basis/1d", "proto_default/1d", "animate_plotly", "animate_gganimate"),
  `related ggplot2 function` =
    c("ggplot", "geom_point", "geom_text", "geom_hex", "NA", "geom- _density & _rect", "geom- _segment & _text", "several protos", "plotly::ggplotly (with animation)", "gganimate::animate"),
  detail =
    c("Also perfroms setup for the tour.", "-", "-", "Heatmap hexegons, for high observation density", "Line segments for the origin, the space where 0 values project to", "1D density with run hash marks underneath, `position = 'stack'` not working with {plotly}.", ".html widget, row numbers added as tooltip on hover. {plotly} doesn't presicly map all {ggplot2} settings; legends, point size and opacity may vary.", "Direction and magnetude of variables to the projection disp~", "Default protos for 2/1D tours", ".gif animation. {gganimate} consumes native ggplots, aestheics should be consistant."), check.names = F
) %>%
  knitr::kable()
```
