---
title: "spacemanip Vignette"
author: "Nicholas Spyrison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette  #.html output error, temp fix doesn't work with html_vingette. worked on laptop
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

[ADD] relationship with tourr and motivation.

# Example 1 - Rotate Flea data

Suppose we want to perform a rotation on one (or more) dimensions on the projection of a higher dimensional (p>3D) space/data. 

Let: 

- n be number of observations (rows)
- p be number of dimensions/numeric variables (columns)
- d be the number of dimensions (linear combinations) of projections from the original p-space. Almost always d=2 for the basis, and d=3 for the manipulation space for our purposes.

In order to rotate a dimension in p-space we need the following args: 

- `basis`, a starting basis, an orthonormal linear combination of p-dimensions, 
- `manip_var`, which dimension(s) to rotate,
- `phi`, angle controlling the magnitude of change of `manip var`
- `theta`, the angle of the change of the `manip var`

For a test data set use flea [n=74, p=6]. In order to get to a manipulation space (`manip_space`) use `create_manip_space(basis, manip_var)`. Below, we will generate a random basis [p, d=2] via `basis_random(p)` and arbitrarily let the `manip_var` be 3.

```{r}
#devtools::install_github("nspyrison/spacemanip")
library(spacemanip)

data <- flea[,1:6]
p <- ncol(data)

this_manip_space <- create_manip_space(basis = basis_random(p = p), manip_var = 3)
```

We then rotate the 3rd variable by an arbitrarily desired magnitude, `phi`, at an arbitrarily desired magnitude, `theta`. The [p, d+1=3] rotated space, (`r_space`) is returned. This is the the linear combination for the xyz contributions in p-space for this rotation.

```{r}
radial_manip(manip_space = this_manip_space, phi = pi/3, theta = pi/4)
```

For illustration plot the projection of the data onto this new `r_space` and plot the xy contributions. Do this several times while varying the magnitude of of `phi` an angle controlling the magnitude of manipulation of the `manip_var`. At a constant `theta` angle of rotation in coordinate space.

```{r}
#knitr doesn't like par(mfrow), maybe ggplot2 has alternative?
for (i in seq(1/5*pi, 2/5*pi, pi/20)) {
  this_r_space <- horizontal_manip(manip_space = this_manip_space, phi = i, theta = 1/3*pi)
  proj <- data_proj(data = data, r_space = this_r_space)
  plot(proj[,1], proj[,2], main="Projected data")
  Sys.sleep(time=.5)
  print(i)
}
```

# Example 2 - Climate glyph map


```{r}
library(GGally)
?GGally::glyphs #see example
 nasaLate <- nasa[
   nasa$date >= as.POSIXct("1998-01-01") &
   nasa$lat >= 20 &
   nasa$lat <= 40 &
   nasa$long >= -80 &
   nasa$long <= -60
 , ]
 temp.gly <- glyphs(nasaLate, "long", "day", "lat", "surftemp", height=2.5)
 ggplot2::ggplot(temp.gly, ggplot2::aes(gx, gy, group = gid)) +
   add_ref_lines(temp.gly, color = "grey90") +
   add_ref_boxes(temp.gly, color = "grey90") +
   ggplot2::geom_path() +
   ggplot2::theme_bw() +
   ggplot2::labs(x = "", y = "")
```

# Example 3 - Tour manipulation

```{r}
library(tourr)

data <- flea[,1:6]
p <- ncol(data)
col <- rainbow(length(unique(flea$species)))[as.numeric(as.factor(flea$species))]
pch <- as.numeric(flea$species)+14

holes <- save_history(data, guided_tour(index = holes), max_bases = 15)
holes_end <- matrix(as.numeric(holes[,,dim(holes)[3]]),ncol=2)

m_sp <- create_manip_space(basis = holes_end, manip_var = 3)



```

# Summary

[RECAP WHAT PKG HELPS US DO, nanier for example:]
The tools in `naniar` help us identify where missingness is, while maintaining a tidy workflow. We care about these mechanisms or these patterns because they can help us understand potential mechanisms, such as equipment failures, and then identify possible solutions based upon this evidence.

# Future development

[further integration with tourr needed?]

# Thank you

Formost, thanks to Di Cook for direction, inspiration, and the ground work underpinning this work. Her ability for conceptualizing higher dimensions and visualizing data in general is unparallel.
Thanks to Ursla Laa for use cases and prompts to keep me focused.

# References
<!-- Monash prefers APA -->
<!-- citation(package = "tourr") -->
<!-- citation(package = "GGally") -->

- Cook, D., & Buja, A. (1997). Manual Controls for High-Dimensional Data Projections. Journal of Computational and Graphical Statistics, 6(4), 464–480. https://doi.org/10.2307/1390747
- Dianne Cook and Hadley Wickham (NA). tourr: Implement Tour Methods in R Code. Rpackage version 0.5.5.9000. https://github.com/ggobi/tourr
- Barret Schloerke, Jason Crowley, Di Cook, Francois Briatte, Moritz Marbach,
Edwin Thoen, Amos Elberg and Joseph Larmarange (2017). GGally: Extension to
'ggplot2'. R package version 1.3.2. https://CRAN.R-project.org/package=GGally


