---
title: "Animated pdf example"
author: "Nicholas Spyrison"
date: "6/2021"
output:
  #html_document:
  pdf_document:
  #bookdown::pdf_document2:
    toc: true
header-includes:
  - \usepackage{animate}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE
)
```

# Setup and gotchas

Keep an eye on YAML header and chunk options.

1) YAML must load thelatex package `animate`, _ie._ `header-includes:  - \usepackage{animate}`
2) Chunk option of the animation needs:  `fig.show='animate', interval=1/fps`, and sometimes `animation.hook='gifski'`
3) For use with `{rticles}` templates and other local templates, you may need to add the following to the `template.tex` preamble:

<!-- Will be the 3 lines with `$`, without the r chunk, directly into the *.tex. -->
```{r, eval=F,echo=T}
$if(highlighting-macros)$
$highlighting-macros$
$endif$
```

# base R

```{r pacmanExample, animation.hook='gifski', fig.show='animate', interval=.2}
for (i in 1:3){
  pie(c(i %% 3, 6), col = c('red', 'yellow'), labels = NA)
}
```



# {gganimate}
```{r gganimate, fig.show='animate', interval=.2}
library("gganimate")

ggplot(datasets::airquality, aes(Day, Temp, group = Month)) +
  geom_line() +
  transition_reveal(Month)

message("Seems to produce many intermediate frames, `interval` may not be be stright forward with gganimate.")
```


# {spinifex}, ggproto API 
```{r spinifex, fig.show='animate', interval=.2,      eval = F}
##TODO, still implementing spinifex ggproto api.

library("spinifex")

#?animate_gganimate_knit2pdf
dat <- scale_sd(tourr::flea[, 1:6])
clas <- tourr::flea[, 7]
bas <- basis_pca(dat)
mv <- manip_var_of(bas)
mt <- manual_tour(bas, manip_var = mv, angle = .1)

ggtour <- ggplot_tour(mt_array, dat) +
  ggproto_basis_axes() +
  ggproto_data_background() +
  ggproto_data_points(aes_args = list(color = clas, shape = clas),
                      identity_args = list(size= 1.5, alpha = .7))

animate_gganimate_knit2pdf(ggtour)
```

<!-- # {spinifex}, OLD API  -->
<!-- ```{r spinifex, fig.show='animate', interval=.2} -->
<!-- libray("spinifex") -->

<!-- #?play_tour_path -->
<!-- dat_std <- scale_sd(flea[, 1:6]) -->
<!-- clas <- flea$species -->
<!-- bas <- basis_pca(dat_std) -->
<!-- ## Not run: -->
<!-- ## Tour history from tourr::save_history -->
<!-- g_path <- tourr::save_history(dat_std, tour_path = tourr::grand_tour(), max = 3) -->
<!-- ## Recreate play_tour_path(render_gganimate) -->
<!-- gg <- play_tour_path(tour_path = g_path, data = dat_std, render_type = render_) -->
<!-- gg + gganimate::transition_states(frame, transition_length = 0L) -->
<!-- #gganimate::knit_print.gganim(gga) -->

<!-- message("Only works with an implicit call, but not with animate(), difference with print.gganim() and knit_print.gganim()??") -->
<!-- ``` -->


<!-- # tourr -- only 2 frames? -->
<!-- Only ever gets to the first 2 frames... -->
<!-- ```{r Tourr, animation.hook='gifski', fig.show='animate', interval=.2} -->
<!-- require(tourr) -->
<!-- if(interactive() == T) -->
<!--   ?save_history -->
<!-- t1 <- save_history(flea[, 1:6], max = 4) -->
<!-- animate_xy(flea[, 1:6], planned_tour(t1)) -->

<!-- message("why only 2 frames!?") -->
<!-- ``` -->